Practical Optional Types for Clojure

### 2015-02-28 formal-model.tex
experience
done mm formal

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    %{\TDo}
    %{\TClass}
    %{\TIf}
    {\TAbs}
    {\TSubsume}
    \begin{array}{c}
      {\TConst}\\\\
      {\TKw}\\\\
      {\TClass}
    \end{array}
    \begin{array}{c}
      {\TTrue}\\\\
      {\TFalse}\\\\
      {\TNil}
    \end{array}

@@

The main interest for the typing rule is \isacompareliteral{} 
(\figref{main:figure:mmsyntax}).

@@

To ease presentation, we present \emph{immutable}
multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. \egref{example:rep} is now written
\begin{minted}{clojure}
(let [rep (defmulti [Any -> String] class)
      rep (defmethod rep Keyword [x] (str (name x)))
      rep (defmethod rep Number [x] (str (inc x)))]
  (rep :a)) ;=> "a"
\end{minted}

Multimethod semantics are in \figref{main:figure:mmsyntax}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (\figref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(\figref{main:figure:mmsyntax}).

>>>>

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    %{\TDo}
    %{\TClass}
    %{\TIf}
    {\TAbs}
    \begin{array}{c}
      {\TSubsume}\\\\
      {\TNum}
    \end{array}
    \begin{array}{c}
      {\TConst}\\\\
      {\TKw}\\\\
      {\TClass}
    \end{array}
    \begin{array}{c}
      {\TTrue}\\\\
      {\TFalse}\\\\
      {\TNil}
    \end{array}

@@

The main interest for the T-IsA typing rule is the \isacompareliteral{} 
metafunction
(\figref{main:figure:mmsyntax}), used to calculate the propositions for
\isaliteral{} expressions.

@@

To ease presentation, we present \emph{immutable}
multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. \egref{example:rep} is now written
\begin{minted}{clojure}
(let [path (defmulti [Any -> (U nil String)] class)]
  (let [path (defmethod path String [x] x)]
    (let [path (defmethod path File [^File x] 
                 (.getPath x))]
      (path "dir/a")))) ;=> "a"
\end{minted}
where subsequent bindings in the binder is sugar for nested let expressions.

The syntax for multimethods is given in \figref{main:figure:mmsyntax}.
{\MultiFntype {\s{}} {\sp{}}} is the type of multimethods with \emph{interface type}
{\s{}} and \emph{dispatch function type} {\sp{}}.
The expression {\createmultiexp {\s{}} {\e{}}} expression creates a multimethod
with interface type \s{} and dispatch function \e{}.
The expression {\extendmultiexp {\e{m}} {\e{v}}{\e{f}}}
takes multimethod \e{m} and returns a new multimethod with an extended dispatch
table from dispatch value {\e{v}} to {\e{f}}.
The value {\multi {\v{}} {\disptable{}}} is the runtime value of a multimethod
with dispatch function {\v{}} and dispatch table {\disptable{}}.

The typing rules are in \figref{main:figure:mmsyntax}.
The T-DefMulti rule checks to ensure the type of dispatch function
is a function that has at least as permissive a parameter type
as the interface type.

For example, let us check the \defmultiliteral{} from our translation of \egref{example:rep}
using T-DefMulti.

$$
\judgement{\propenv{}}
{\hastype
{\createmultiexp 
      {\s{}}
      {\classconst{}}}
  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
where \s{}  = {\ArrowOne {\x{}} {\Top{}} {\t{}} {\filterset {\topprop{}} {\topprop{}}} {\emptyobject{}}}
  and \sp{} = {\ArrowOne {\x{}} {\Top{}} {\Union{\Nil}{\Class}} {\filterset {\topprop{}} {\topprop{}}} {\path{\classpe{}}{\x{}}}}.
  Since the parameter types agree, this is well-typed.

The T-DefMethod rule is carefully constructed to ensure we have a syntactic
lambda expression as the right-most subexpression.
This way we can manually check the body of the lambda under an extended
environment like we sketched in \egref{example:incmap}.
We use \isacompareliteral{} to infer the proposition for this method.

Let us continue with the next line of the translation of \egref{example:rep}.
From our previous line we have \propenvp{} = \propenv{},{\isprop{\MultiFntype {\s{}}{\sp{}}}{path}}.
$$
\judgement{\propenvp{}}
{\hastype
  {\extendmultiexp {prop} {\String}
                   {\abs {\x{}} {\Top{}} {\x{}}}}
  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
We know \emph{prop} is a multimethod by \propenvp{}, so now we check the body.
$$
\judgement{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
{\hastype
  {\x{}}
  {\String}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
This is checked by T-Local since {\inpropenv{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
The new proposition {\isprop{\String}{\x{}}} was inferred by 
$$
  \isacompare{\Top{}}{\path{\classpe{}}{\x{}}}{\Value{\File{}}}
             {\filterset{\isprop{\String}{\x{}}}
                        {\notprop{\String}{\x{}}}}
$$

The next line is similarly. The final line is checked by T-App because
subtyping for multimethods (\figref{main:figure:mmsyntax}), specifically S-PMultiFn
says that {\MultiFntype {\s{}}{\sp{}}} can be upcast to its interface type {\s{}}.
Since {\s{}} has parameter type \Top{}, this holds trivially.

Multimethod semantics are in \figref{main:figure:mmsyntax}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
and uses \getmethodliteral{} (which models Clojure's \clj{get-method}) 
to obtain a method. The call to \getmethodliteral{} only returns a value
if there is exactly one method such that the corresponding dispatch value
is compatible with the current result of the dispatch function by \isaopsemliteral{}.
Finally we return the result of applying the obtained method and the original evaluated argument.
